---
title: "Washington Marijuana Business Analysis"
author: "Ed Sarausad"
date: "January 22, 2016"
output: html_document
---

```{r, echo=FALSE}

library(broom) # used to neatly pass model results (model chaining/iteration)
library(forecast) #consider using plot.ly for interactive forecast
library(plyr)
library(dummies)
library(nortest)
library(glmnet) # http://web.stanford.edu/~hastie/glmnet/glmnet_alpha.html
library(caret)
library(Amelia) # missing value viz
library(stats)

```

# DATASET/QUESTIONS 

The datasets for sales, violations, and businesses (applicants for marijuana business licenses)

```{r, echo=FALSE}
sales <- read.csv("https://www.dropbox.com/s/yxgij2mtsou0lf0/mj-sales-transformed.csv?dl=1")
violations <- read.csv("https://www.dropbox.com/s/euwbe6mciu9oaw9/mj-violations-transformed.csv?dl=1")
applicants <- read.csv("https://www.dropbox.com/s/hw3ci1l8w55wwxw/all_applicants.csv?dl=1")

```

Research questions:
(1) What is relationships between sales and violations?
(2) What are the predictors of violations?
(3) How can we segment the businesses (identified in applicants data) by risk (likelihood of violating)? 

# DATA EXPLORATION

Let's see what's missing 

```{r}
missmap(applicants, main = "Applicants: Missing values vs observed")
missmap(sales, main = "Sales: Missing values vs observed")
missmap(violations, main = "Violations: Missing values vs observed")

```

Looking for anomalies in the data

```{r}
summary(applicants$ReasonAction) # most applications PENDING, not APPROVED
summary(applicants$State.1) # 13 out of WA state businesses
summary(applicants$type) # PROCESSOR and RETAILER most prevalent

```

Format currency fields and add a tax rate, to see if tax rate is a predictor

```{r}
# TODO some cells have () (negative) and -
sales$Total_Sales <- as.numeric(gsub("\\$","", gsub("\\,","", sales$Total_Sales))) 
# TODO some cells have () (negative) and -
sales$Excise_Tax_Due <- as.numeric(gsub("\\$","", gsub("\\,","", sales$Excise_Tax_Due))) 
# calculate a tax rate (not always 25%)
sales$tax_rate <- sales$Excise_Tax_Due / sales$Total_Sales

```

Aggregate sales and violations by *License_Number* which uniquely identified a marijuana business

```{r eval=FALSE}
# use applicants and aggregate sales
sales_only <- sales[,c(2,4,5,8)]
# BUG BUG gives an error when run through knitr, but works fine in console!
aggsales <- aggregate(sales_only, by=list(sales_only$License_Number), FUN=sum)
# BUG BUG aggsales$License_Number not correct!
# why does the following line put join column in Group.1?
aggsales[aggsales$Group.1==413287,]
# HACK *replacing* value in License_Number with Group.1
aggsales$License_Number <- aggsales$Group.1

# use applicants and aggregate sales and violations
# aggviolations <- count(violations, c('License_Number'))
# NOTE: some businesses use same license to operate different business types (processor, retailer) with diff't tax treatment
# rename count frequency to something meaningful
names(aggviolations)[names(aggviolations)=='freq'] <- "violation_count"

# do a left join to merge aggregate sales and violations to license number
applicants <- merge(applicants, aggsales, by='License_Number', all.x=TRUE) #left join
applicants <- merge(applicants, aggviolations, by='License_Number', all.x=TRUE) #left join

# check the merge!
str(applicants) # Total_Sales and violation_count should be added
```

Handle missing values based on business assumptions

```{r eval=FALSE}
# not evaluating because of knitr/aggregate error

# state missing values assumptions:
#  sales - business has not made money or is not reporting, so setting NAs in Total_Sales, Excise_Tax_Due and tax_rate to 0
#  violations - innocence assumed, so setting to 0
applicants$Total_Sales[is.na(applicants$Total_Sales)] <- 0
applicants$Excise_Tax_Due[is.na(applicants$Excise_Tax_Due)] <- 0
applicants$tax_rate[is.na(applicants$tax_rate)] <- 0
applicants$violation_count[is.na(applicants$violation_count)] <- 0
```


Now looking at normality of businesses **Total_Sales** to see what we techniques we use to use statistical inference

```{r eval=FALSE}
plot(density(applicants$Total_Sales))

```

# FEATURE ENGINEERING


We're interested in whether the type of business (denoted by Privilege Description **PrivDesc**) is a predictor

```{r eval=FALSE}
applicants <- dummy.data.frame(names='type', applicants, sep=":")
# TODO add variables for high-risk counties or cities
applicants$violator <- applicants$violation_count > 0
```


```{r eval=FALSE}
applicants$log_Total_Sales <- log(applicants$Total_Sales)
applicants$log_Excise_Tax_Due <- log(applicants$Excise_Tax_Due)
```

```{r}
# we load transformed applicants data to be on same page after all transformations (due to knitr/aggregate conflict)
applicants <- read.csv("https://www.dropbox.com/s/s168am9a1iknwi8/applicants_transformed.csv?dl=1")

```

# MODEL SELECTION

We try logistic regression to predictor whether a business is a **violator**

```{r}

# logit regression for predicting a violator
train <- applicants[1:4130,]
test <- applicants[4131:5903,]
model <- glm(violator ~ Total_Sales + Excise_Tax_Due + PrivDesc,family=binomial(link='logit'),data=train)
summary(model) # significant: sales, tax, retailer type

model <- glm(violator ~ Total_Sales + Excise_Tax_Due + ReasonAction,family=binomial(link='logit'),data=train)
summary(model) # significant: pending, withdrawn

model <- glm(violator ~ Total_Sales + Excise_Tax_Due + PrivilegeStatus,family=binomial(link='logit'),data=train)
summary(model) # slightly significant: issued

model <- glm(violator ~ Total_Sales + Excise_Tax_Due + tax_rate,family=binomial(link='logit'),data=train)
summary(model) # 
```

# MODEL EVALUATION

ROC (Receiver Operating Characteristic) used to evaluate model performance

